---
title: "Missing data and temporal models"
author: "Monica Alexander"
date: "March 15 2022"
output: 
  pdf_document
---

# Child mortality in Sri Lanka

In this lab you will be fitting a couple of different models to the data about child mortality in Sri Lanka, which was used in the lecture. Here's the data and the plot from the lecture:

```{r}
Sys.setenv(BINPREF = 'C:/rtools40/mingw64/bin/')

library(tidyverse)
library(here)
library(rstan)
library(tidybayes)
lka <- read_csv("lka.csv")
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```

# Fitting a linear model 

Let's firstly fit a linear model in time to these data. Here's the code to do this:

```{r}
observed_years <- lka$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)
stan_data <- list(y = lka$logit_ratio, 
                  year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = mean(years), se = lka$se)
```


```{r}
# mod <- stan(data = stan_data,
#              file = "lka_linear_me.stan")
# 
# saveRDS(mod, "mod.rds")
mod <- readRDS("mod.rds")
```

Extract the results:

```{r}
res <- mod %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year = years[t])
```


Plot the results:

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res, aes(year, .value)) + 
  geom_ribbon(data = res, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Linear fit shown in black")
```

## Question 1

Project the linear model above out to 2022 by adding a `generated quantities` block in Stan (do the projections based on the expected value $\mu$). Plot the resulting projections on a graph similar to that above. 


             
```{r}
stan_data <- list(y = lka$logit_ratio, 
                  year_i = observed_years - years[1]+1, 
                  T = nyears, 
                  years = years, 
                  N = length(observed_years), 
                  mid_year = mean(years), 
                  se = lka$se,
                  P = nyears + 8,
                  new_years = c(years,seq(2015, 2022)))

# mod2 <- stan(data = stan_data,
#              file = "lka_proj2.stan")
# 
# saveRDS(mod2, "mod2.rds")
mod2 <- readRDS("mod2.rds")
```



```{r}
res2 <- mod2 %>% 
  gather_draws(y_rep[t]) %>%
  median_qi() %>%
  mutate(year = t + 1951)
  
```


# Random walks


## Question 2

Code up and estimate a first order random walk model to fit to the Sri Lankan data, taking into account measurement error, and project out to 2022. 

```{r}
stan_data <- list(y = lka$logit_ratio, 
                  year_i = observed_years - years[1]+1, 
                  T = nyears, 
                  years = years, 
                  N = length(observed_years), 
                  mid_year = mean(years), 
                  se = lka$se,
                  P = nyears + 8,
                  new_years = c(years,seq(2015, 2022))
                  )

# mod_rw1 <- stan(data = stan_data,
#              file = "lka_previous_1.stan")
# 
# saveRDS(mod_rw1, "mod_rw1.rds")
mod_rw1 <- readRDS("mod_rw1.rds")
```


```{r}
res_rw1 <- mod_rw1 %>% 
  gather_draws(y_rep[t]) %>%
  median_qi() %>%
  mutate(year = t + 1951)
```



## Question 3

Now alter your model above to estimate and project a second-order random walk model (RW2). 

```{r}
stan_data <- list(y = lka$logit_ratio, 
                  year_i = observed_years - years[1]+1, 
                  T = nyears, 
                  years = years, 
                  N = length(observed_years), 
                  mid_year = mean(years), 
                  se = lka$se,
                  P = nyears + 8,
                  new_years = c(years,seq(2015, 2022))
                  )

# mod_rw2 <- stan(data = stan_data,
#              file = "lka_rw2.stan")
# 
# saveRDS(mod_rw2, "mod_rw2.rds")
mod_rw2 <- readRDS("mod_rw2.rds")

res_rw2 <- mod_rw2 %>% 
  gather_draws(y_rep[t]) %>%
  median_qi() %>%
  mutate(year = t + 1951)


```


## Question 4

Run the first order and second order random walk models, including projections out to 2022. Compare these estimates with the linear fit by plotting everything on the same graph. 

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res_rw1, aes(year, .value)) + 
  geom_ribbon(data = res_rw1, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_rw2, aes(year, .value), color = 'red') + 
  geom_ribbon(data = res_rw2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = 'red')+
  geom_line(data = res2, aes(year, .value), color = 'blue') + 
  geom_ribbon(data = res2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "blue")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Random Walk fit shown in black")
```

## Question 5

Rerun the RW2 model excluding the VR data. Briefly comment on the differences between the two data situations. 

```{r}
lka1 <- lka %>%
  filter(source != 'VR')

observed_years <- lka1$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

stan_data <- list(y = lka1$logit_ratio, 
                  year_i = observed_years - years[1]+1, 
                  T = nyears, 
                  years = years, 
                  N = length(observed_years), 
                  mid_year = as.integer(mean(years)), 
                  se = lka1$se,
                  P = nyears + 8,
                  new_years = c(years,seq(2015, 2022)))

# mod_rw3 <- stan(data = stan_data,
#              file = "lka_rw2.stan")
# 
# saveRDS(mod_rw3, "mod_rw3.rds")
mod_rw3 <- readRDS("mod_rw3.rds")
```



```{r}
res_rw3 <- mod_rw3 %>% 
  gather_draws(y_rep[t]) %>%
  median_qi() %>%
  mutate(year = t + 1951)
  
```


```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res_rw3, aes(year, .value)) + 
  geom_ribbon(data = res_rw3, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_rw2, aes(year, .value), color = 'red') + 
  geom_ribbon(data = res_rw2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = 'red')+
  # geom_line(data = res2, aes(year, .value), color = 'blue') + 
  # geom_ribbon(data = res2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "blue")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Random Walk fit shown in black")
```
## Question 6

Briefly comment on which model you think is most appropriate, or an alternative model that would be more appropriate in this context. 

I think the second RW2 without VR is the most appropriate since it has the closest ci's for projections.

Ideally, use loo to check, but I am running out of time.

```{r}
# to use
loglik2 <- rstan::extract(mod_rw2)[["log_lik"]]
loglik3 <- rstan::extract(mod_rw3)[["log_lik"]]

loo2 = loo(loglik2, save_psis = T )
loo3 = loo(loglik3, save_psis = T )
```



Here is my model, but it is overfitted.
```{r}
observed_years <- lka$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

stan_data <- list(y = lka$logit_ratio, 
                  year_i = observed_years - years[1]+1, 
                  T = nyears, 
                  years = years, 
                  N = length(observed_years), 
                  mid_year = mean(years), 
                  se = lka$se,
                  P = nyears + 8,
                  new_years = c(years,seq(2015, 2022))
                  )

# my_mod <- stan(data = stan_data,
#               file = "rw_back.stan")
```


```{r}
#saveRDS(my_mod, "my_mod.rds")
my_mod <- readRDS("my_mod.rds")
```


```{r}
my_res <- my_mod %>% 
  gather_draws(y_rep[t]) %>%
  median_qi() %>%
  mutate(year = t + 1951)
  
```
I think its overfitting, its not nice:(

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = my_res, aes(year, .value)) + 
  geom_ribbon(data = my_res, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  # geom_line(data = res2, aes(year, .value), color = 'blue') + 
  # geom_ribbon(data = res2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "blue")+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio")
```

